%Background
\textbf{Background}

Synthetic lethal genetic interactions are re-emerging the post-genomics era due to their potential for precision medicine against cancers. Synthetic lethality exploits functional redundancy with genes disrupted in cancers (including tumour suppressors) to develop specific treatments. E-cadherin, encoded by \textit{CDH1}, is a tumour supressor gene with loss of function in breast and btomach cancers. Experimental screens have identified candidate synthetic lethal interactions for drug target triage which can be further supported with bioinformatics analysis. Furthermore, gene expression data is amenable to Investigations of the pathway composition and structure of synthetic lethal candidates. 

%Methods \ Approach
\textbf{Approach}

A computational methodology, the Synthetic Lethal Prediction Tool (\acrshort{SLIPT}) was developed to detect synthetic lethal interactions in gene expression data. This methodology was demonstrated on interactions with \textit{CDH1} in breast and stomach cancer data from The Cancer Genome Atlas (\acrshort{TCGA}) project. Synthetic lethal genes and pathways were further investigated with unsupervised clustering, gene set over-representation analysis, metagenes, and permutation resampling. In particular, analyses focused on comparing \acrshort{SLIPT} gene candidates to an experimental \acrshort{siRNA} screen \citet{Telford2015}. Graph theory methods were also applied to the most supported pathways to test for pathway structure among between synthetic lethal candidates. Simulation and modelling was used to assess the statistical performance of \acrshort{SLIPT}, including simulated data with correlation structures derived from graph stuctures.

%Results \ Findings
\textbf{Findings}

%chapt 3
Many candidate synthetic lethal partners of \textit{CDH1} were detected in both \acrshort{TCGA} breast cancer. These genes clustered into several distinct groups, with distinct biological functions and elevated expression in different clinical subtypes. While the number of genes detected by both approaches was not significant, these contained significantly enriched pathways. In particular $G_{\alpha i}$ signalling, cytoplasmic microfibres, and extracelluar fibrin clotting were robustly supported by both approaches which is consistent with the known cytoskeletal and cell signalling roles of E-cadherin and validation of \acrshort{GPCR}s perfomed by \citet{Telford2015}. Many of these pathways were replicated in stomach cancer. The pathways supported only by \acrshort{SLIPT} included regulation of immune signalling and translational elongation which were unlikely to be detected in an isogenic cell line model but are are still candidates for further investigation. 

%chapt 4
Synthetic lethal candidates detected by \acrshort{SLIPT} and \acrshort{siRNA} were compared within graph structures of the candidate synthetic lethal pathways. These genes did not differ by network measures of importance or connectivity in the pathway. There was also little support for \acrshort{SLIPT} gene candidates being upstream or downstream of \acrshort{siRNA} gene candidates within a pathway, consistently across pathways. 

%chapt 5

A model of synthetic lethality used to simulate gene expression datasets with synthetic lethal partners of a gene. The \acrshort{SLIPT} methdology had high statistical performance detecting few synthetic lethal partners, although this diminishes with more synthetic lethal parners or lower sample size. The \acrshort{SLIPT} methdology performs better than Pearson's correlation or the $\chi^2$-test. In particular, it performs well with high specificity for dtasets containing thousands of genes or genes positively correlated with the query gene (as expected in human expression data). It was also robust across correlation structures, including those derived from complex pathway structures and often distinguished synthetic lethal genes from those positively or negatively correlated with them. Therefore \acrshort{SLIPT} is appropriate to identify synthetic lethal genes within pathways and use candidate synthetic lethal genes (and their correlates) to identify synthetic lethal pathways.

%Conclusions
\clearpage
\textbf{Summary}

Thus my thesis has developed, evaluated and refined a bioinformatics approach to discovery of synthetic lethal genes solely from gene expression data. This approach has been demonstrated to detect biologically informative and clinically relevant candidate partners for \textit{CDH1} in breast and stomach cancers. These investigations have also involved the development of network analysis and simulation procedures which may be more widely applicable.

\clearpage

\begin{center}
 \textbf{Research Contributions During Candidature}
\end{center}

\textbf{Publications}

\begin{small} \begin{flushleft} Kelly, S. T. and Spencer, H. G. (2017) Population-Genetics Models of Sex-Limited Genomic Imprinting. \textit{Theoretical Population Biology} \textbf{115}:35-44 \doi{10.1016/j.tpb.2017.03.004} \end{flushleft} \end{small}

\textbf{Manuscripts Submitted}

%Several publications have been prepared during the candidature of this thesis, including some findings related to the thesis topic on which we elaborate in more depth here. Please see the original article for the results which have been accepted for publication by peer-review:
\begin{small}
Kelly, S. T., Single, A. B., Telford, B. J., Beetham, H. G, Godwin, T. D., Chen, A., Black, M., A., and Guilford, P. J. (2017) Towards HDGC chemoprevention: vulnerabilities in E-cadherin-negative cells identified by genomic interrogation of isogenic cell lines and whole tumors.  Submitted to \textit{Cancer Prevention Research}.

Kelly, S. T., Chen, A., Guilford, P. J., and Black, M. A. (2017) Synthetic lethal interaction prediction of target pathways in E-cadherin deficient breast cancers. Submitted to \textit{BMC Genomics}.
\end{small}
\iffalse
\textbf{Community Blog Posts}

Black, M. A., Kelly, S. T., and Cadzow, M.
Posted on the \textit{Software Carpentry} website 2016 July 4\textsuperscript{th}: 
``Software Carpentry workshop at the University of Otago, New Zealand''
\url{https://software-carpentry.org/blog/2016/07/otago-workshop.html}

Kelly, S. T., Black, M., A., Bae, S., Hayek, W., and Pawlik, A. Posted on the \textit{Software Carpentry} website 2016 September 28\textsuperscript{th}:  ``Software Carpentry Workshop Attendance: a New Zealand Perspective``
\url{https://software-carpentry.org/blog/2016/09/attendance-nz.html}
\fi

\textbf{Conference Presentations}

\begin{small}
Consortium of Biological Sciences 2017 (Kobe) December TBC

eResearch 2017 (Queenstown) February 20\textsuperscript{th}-22\textsuperscript{nd}
%``Detecting Synthetic Lethality from Cancer Gene Expression: A PhD project on genetic interactions with CDH1 inactivation in TCGA data''

Research Bazaar 2016 (Dunedin) February 2\textsuperscript{nd}-4\textsuperscript{th}

eResearch 2016 (Queenstown) February 9\textsuperscript{th}-11\textsuperscript{th}
%``Sifting the Needles in the Haystack: Permutation Resampling Biological Pathways in Cancer Genomic Interaction Data'' (Supported by REANNZ)

Genetics Otago Symposium 2016 (Dunedin) March 7\textsuperscript{th}-8\textsuperscript{th} 
%``A Bioinformatics approach to Genetic Interactions: Synthetic Lethal Pathways with E-cadherin in Breast Cancer Genomics Data''

DunDead: Zombie Science and Culture Festival 2014 (Dunedin) %Ignite Speaker
August 16\textsuperscript{th}-17\textsuperscript{th}
%``Hidden in Plain Sight - The Genetics of Zombies''

eResearch 2014 (Hamilton) %Ignite Speaker
June 30\textsuperscript{th}-July 2\textsuperscript{nd}
%``Bioinformatic analysis of synthetic lethal genetic interactions in breast cancer''
(Supported by Google)

\end{small}

\textbf{Poster Presentations}

\begin{small}
Next Generation Sequencing Asia 2016 (Singapore) October 11\textsuperscript{th}-12\textsuperscript{th}
%``Bioinformatic Investigations of Synthetic Lethal Interactions with E-cadherin in Breast Cancer''
(Supported by the University of Otago Division of Health Sciences; Maurice and Phyllis Paykel Trust)

Research Bazaar 2015 (Melbourne, Australia) February 16\textsuperscript{th}-18\textsuperscript{th}
%``My digital research toolkit''
(Supported by the New Zealand eScience Infrastructure)

Otago School of Medical Sciences Postgraduate Symposium 2015 (Dunedin, New Zealand) April 28\textsuperscript{th}-29\textsuperscript{th}

QMB Cancer Drugs Satellite 2014 (Queenstown, New Zealand) August 24\textsuperscript{th}-25\textsuperscript{th}
%“Bioinformatics Prioritisation of Synthetic Lethal Targets for Drug Activity Against E-Cadherin Deficient Cancers”

\end{small}

\clearpage
\textbf{Seminar Presentations}

\begin{small}

University of Otago Department of Biochemistry 2017 (Dunedin) November TBC

Tohoku University 2016 (Sendai) November 11\textsuperscript{th}

Okinawa Institute of Science and Technology 2016 (Onna) November 1\textsuperscript{st}

Sokendai Graduate University 2016 (Hayama) October 25\textsuperscript{th}

Tokyo University Institute of Medical Science 2016 (Shirokanedai) October 24\textsuperscript{th} 

National Institute of Genetics 2016 (Mishima) October 21\textsuperscript{st}

RIKEN Division of Genomic Technologies 2016 (Yokohama) October 20\textsuperscript{th}
%``Analysis of Synthetic Lethal Pathways in Breast Cancer: A PhD project on genetic interactions with CDH1 inactivation in TCGA data''

\end{small}




\iffalse
Presentations within University of Otago Departments
Department of Biochemistry 2017 PhD Seminar
Upcoming Seminar (to schedule later upon completion of thesis)
``Synthetic Lethal Interactions in Cancer Genomics Data: Prediction and Simulation''

Department of Biochemistry 2016 Colloquia July 22nd
``Analysis of Synthetic Lethal Pathways in Breast Cancer: A PhD project on genetic interactions with CDH1 inactivation in TCGA data''
``Demonstration of R data visualization packages: vioplotx and heatmap.2x''

Otago Mozilla Study Group 2016 June 9th
``Taking Advantage of Online Communities: Getting the most out of StackExchange and StackOverflow Q&A Sites''

Genetics Otago Journal Club 2016 March 23rd 
Day T, Read AF (2016) Does High-Dose Antimicrobial Chemotherapy Prevent the Evolution of Resistance? PLoS Comp Biol 12(1): e1004689

Otago Mozilla Study Group 2015 October 22nd
``Data is Beautiful: An Introduction to Data Visualization in R''

Department of Biochemistry 2015 Colloquia August 27th
``A Demonstration of Statistical Analysis in R: Antibiotic Resistance in Biofilms'' (collaborator data)

Otago Mozilla Study Group 2015 June 9th
``Literature Tools: Bibliometrics and Reference Management''

Biochemistry Department Journal Club 2015 April 25th
``Quelling the Flames: Order Matters in Cancer''
Ortmann CA, Kent DG, Nangalia J, et al. (2015) Effect of mutation order on myeloproliferative neoplasms. N Engl J Med 372, 601-612.

Genetics Otago Journal Club 2014 September 3rd
Peris et al. (2014) Population structure and reticulate evolution of Saccharomyces eubayanus and its lager-brewing hybrids. Molecular Ecology 23(8):2031-2045

\fi




\textbf{Software Packages}

Software packages in the R language have been released. Please refer to the appropriate GitHub repository for more information (including documentation, vignettes, and installation instructions), on the following account: \url{https://github.com/TomKellyGenetics}

\begin{small}

\begin{itemize}
 \item \href{https://github.com/TomKellyGenetics/slipt}{\texttt{slipt}} to accompany the synthetic lethal publication above and release SLIPT (Synthetic Lethal Interaction Prediction Tool)
 \item \href{https://github.com/TomKellyGenetics/vioplotx}{\texttt{vioplotx}} to provide enhanced violin plots
 \item \href{https://github.com/TomKellyGenetics/heatmap.2x}{\texttt{heatmap.2x}} to provide annotated heatmaps
 \item \href{https://github.com/TomKellyGenetics/igraph.extensions}{\texttt{igraph.extensions}} metapackage for the packages for iGraph objects:
 \begin{itemize}
  \item \href{https://github.com/TomKellyGenetics/plot.igraph}{\texttt{plot.igraph}} to provide plotting for directed graphs
  \item \href{https://github.com/TomKellyGenetics/info.centrality}{\texttt{info.centrality}} to compute network analysis metrics
  \item \href{https://github.com/TomKellyGenetics/pathway.structure.permutation}{\texttt{pathway\-.structure\-.permutation}} for resampling within pathways
  \item \href{https://github.com/TomKellyGenetics/graphsim}{\texttt{graphsim}} to simulate expression (\texttt{mvtnorm}) from pathway structures
 \end{itemize}
\end{itemize}

\end{small}

The \href{https://github.com/TomKellyGenetics/slipt-app}{\texttt{slipt-app}} GitHub repository also hosts an application for \gls{SLIPT} developed in the R \texttt{shiny} environment as part of a related project. There is also a digital copy of this thesis, including high resolution full-colour figures, hosted at:

\url{https://github.com/TomKellyGenetics/thesis/blob/master/thesis.pdf}

\iffalse
\href{https://github.com/TomKellyGenetics/slipt}{\texttt{slipt}} is the Synthetic Lethal interaction Prediction Tool, released to accompany the synthetic lethal publication above. \href{https://github.com/TomKellyGenetics/slipt-app}{\texttt{slipt-app}} contains an application developed in the R \texttt{shiny} environment as part of a related project.

Several plotting functions were customised for the Figures in this thesis (and the above publications), notably \href{https://github.com/TomKellyGenetics/heatmap.2x}{\texttt{heatmap.2x}} and \href{https://github.com/TomKellyGenetics/vioplotx}{\texttt{vioplotx}} have been prepared largely for use during this project but are also documented and available to other R users. These are enhancements to the CRAN \texttt{gplots} and \texttt{vioplot} packages respectively and are intended be user-friendly for those familiar with \texttt{heatmap.2} or \texttt{vioplot} and \texttt{boxplot} (base R) functions. These are backwards compatible with these functions, taking similar imputs as demonstrated in the appropriate vignettes.

The use of iGraph (the R \texttt{igraph} package) operations of graph-network structure in the analysis and simulations of pathways involved several original or customised functions to manipuate or plot \texttt{igraph} objects and adjacency matrices. These can be install separately from their respective repositories of with the metapackage: \href{https://github.com/TomKellyGenetics/igraph.extensions}{\texttt{igraph.extensions}}. \href{https://github.com/TomKellyGenetics/plot.igraph}{\texttt{plot.igraph}} enables plotting graph networks with customised inhibitor arrow and node colours.\href{https://github.com/TomKellyGenetics/info.centrality}{\texttt{info.centrality}} enables the calculation of additional node and network centrality metrics not available in the \texttt{igraph} package. \href{https://github.com/TomKellyGenetics/pathway.structure.permutation}{\texttt{pathway\-.structure\-.permutation}} enables testing of related states or node groups in a network by directionality of shortest paths. The \href{https://github.com/TomKellyGenetics/graphsim}{\texttt{graphsim}} package has been set up to simulate a multi-variate normal gene expression dataset with \texttt{mvtnorm} while deriving the correlation structure, $\Sigma$, from a graph structure. Note that these require various packages for graph theory, statistics and matrix operations and these will be installed as dependencies.
\fi

